Predicting Solid State Material Platforms for Quantum Technologies

### 2021-12-10 paper.tex
Updates from Overleaf

%We have developed data extraction tools for numerous material science databases, and constructed over $4800$ physics-informed features for a dataset consisting of more than $25000$ materials. Furthermore, we have developed and implemented three data mining approaches, termed \textit{the Ferrenti approach}, \textit{the augmented Ferrenti approach} and \textit{the insightful approach} for defining three distinct training sets for the supervised machine learning algorithms logistic regression, decision tree, random forest and gradient boost to be trained on.
Data extraction tools have been developed for a number of materials science databases resulting in a dataset consisting of over $25'000$ materials. The information from the databases was converted into nearly $5'000$ physics-informed features. Furthermore, we have developed and implemented three different approaches for the data mining process: \textit{The Ferrenti approach}, \textit{the augmented Ferrenti approach}, and \textit{the intuitive approach}. 
%Rename insightful approach! To what? 
%Options include stringent, strict, cherry-picked, informed and intuitive. 
The three approaches define distinct training sets for the supervised machine learning methods logistic regression, decision trees, random forests and gradient boosting.

@@

After compiling a dataset from the procedure outlined above, we face an obstacle in terms of defining a training set for the machine learning algorithms. This is challenging for two reasons. 
Firstly, although defects like the NV center in diamond and emitters in SiC are becoming
increasingly well studied, few candidate materials are known. An additional consideration
is that the physical mechanisms promoting favorable defect properties are not fully understood. Secondly, defining materials as unsuitable candidates for QT has proven to be surprisingly intricate. 
%This is not only challenging due to the lack of known suitable candidates (1),  but also due to the intricacy of defining materials as unsuitable candidates (0). 
We follow three separate procedures to define a training set consisting of materials that are labeled as both suitable and unsuitable candidates for QT.

@@

We then label \textbf{unsuitable} candidates according to the reverse requirements of the above; as materials in the ICSD database that crystallize in polar space groups, are calculated 
to be magnetic and have a band gap larger than $0.1$~eV (to exclude metals but include lower-band gap semiconductors) as calculated by DFT using the PBE functional.
The resulting complete dataset contains materials where $56 \ \%$ are labeled as unsuitable and $44 \ \%$ as suitable for QT applications.

\subsubsection*{The augmented Ferrenti approach}
In the second approach, we adjust the first approach in order to improve the dataset. This approach is therefore named \emph{the augmented Ferrenti approach}.

@@

For \textbf{unsuitable} candidates, we implement the same strategy as for the non-augmented Ferrenti approach. The result is a somewhat unbalanced dataset, with up to $75 \ \%$ of the 
materials found in the suitable group. However, the dataset is $78 \ \%$ larger than for the Ferrenti approach.

\subsubsection*{The intuitive approach}
The third approach differs substantially from the first two approaches in terms of labeling, as we apply knowledge from the field (see for instance Refs.~\cite{Atatuere2018,Zhang2020,Son2020,Toth2019,Bathen2021} for an overview)   
to pick our promising material host candidates. 
In other words, we restrict ourselves to materials where quantum compatible properties have been either experimentally demonstrated or theoretically predicted. 
We therefore name this scheme \emph{the intuitive approach}. Due to the concern of having a dataset that is too narrow, we will also include some  materials that are promising in the sense that they have suitable properties for accommodating  deep defects that potentially can exhibit quantum effects - even though such effects may not yet have been shown. 
Materials where properties like single-photon emission and spin manipulation are attributed to quantum dots (formed by self-assembly or lithographic structuring) will also be included in this context.  
The approach to pick \textbf{suitable} candidates can be summed up as;   
\begin{enumerate}
    \item Pick candidates that match the formulas in Table~\ref{tab:qt-materials}, or the formulas ZnSe, AlP, GaP, AlAs, ZnTe, CdS \cite{Weber2010} or SiGe \cite{Hardy2019}. 
    \item Make sure that the candidates are present in the ICSD database.  
    \item Perform a manual screening for correct structures. 
\end{enumerate}

@@

Since the dataset constituting suitable candidates in the third approach contains few entries, we add $400$ \emph{unsuitable} candidates. These are picked at random from the pool of unsuitable candidates from the two previous approaches, in addition to those that were marked as unsuitable through the manual screening process.

@@

To visualize the complexity of the training sets, we have found the two largest eigenvalues of the covariance matrix of the initial data from the Materials Project, and transformed the training sets according to the corresponding two eigenvectors. The resulting scatter plots are found in Figure~\ref{fig:2dscatterplotpca}. 
Green squares display suitable candidates and red triangles represent unsuitable candidates. 
Due to the complexity of reducing the large amount of features down to only two, both suitable and unsuitable candidates for the Ferrenti and the augmented Ferrenti approaches are largely overlapping. 
Hence, using these approaches for predicting QT material hosts could prove challenging for any model that would try to glean a clear-cut boundary between materials that are and are not suitable for QT. 
%Nonetheless, we note that the unsuitable candidates appear to be restricted to a certain parameter space while the suitable candidates extend much further.  
Therefore, we can expect that the two Ferrenti approaches would need supplementary dimensions for further distinguishing between the materials in the two categories (suitable and unsuitable), or that it could prove challenging to find a generalized model based on the available datasets.

@@

It is important to note that the constructed datasets encompass compounds formed by a plethora of combinations of surfaces, interfaces, nanostructures, compositions and structures. We note that this complexity is not necessarily reflected in the descriptors. 
%Additionally, we acknowledge that many compositions deemed as suitable candidates consist of either rare or dangerous elements. By utilizing an enormously large database as Materials Project, we have to account for their ultimate goal - to model all possible materials and their properties.
%The automated process of adding an entry to their database does not necessarily contain all relevant information about a respective material. This is information that needs to be added manually.
Furthermore, we have utilized data obtained from high-throughput density functional theory calculations. Indeed, there are possible errors associated with every step, starting from an initial calculation, adding of data in the database, gathering of data, featurization of data, preprocessing of data, data mining, and finally training a model and making a prediction. Unfortunately, if an error has happened in the first part of the process, the error follows the entire process and will get increasingly harder to detect. Therefore, we are dependent on that the Materials Project has obtained data with high quality, and we note that it is likely that there are some errors present in our data sets.

@@

Taking the four machine learning algorithms outlined above and applying them to the datasets optimized using each of the three approaches yields 12 sets of results for predicting  materials as suitable material hosts for QT. In this section, we present representative ML results based on each of the three data mining approaches.

@@

We also find two compositions with the same chemical formula, the orthorhombic coordinated (mp-$629458$) and the chalcopyrite structured BC$_2$N (mp-$1008523$) where the first structure is in a polar space group while the latter is not. The band gaps are in MP calculated as $1.85$ eV and $1.65$ eV, respectively. BC$_2$N is known as heterodiamond and is a hybrid of diamond and BN. Both structures have, as expected, strong covalent character and have been studied for application as nanostructures \cite{Gao2017}, hydrogen storage \cite{Cai2017} and superhard materials \cite{Li2017, Jiang2020} in ab-initio calculations. Of similar compounds, it has been predicted that the diamond-like structure of BC$_3$N can be a prominent spin qubit material host \cite{Wang2020SpinQB}. By creating a boron (B) vacancy, it will immediately lead to an NV center with similar properties as found in the NV center in diamond. If this is also possible for BC$_2$N, remains to be seen. We note that BC$_3$N is not present in MP, and therefore not present in our dataset.

The compounds InAs (mp-$20305$), CuI (mp-$22895$ and mp-$569346$) and ZnCd$_3$Se$_4$ (mp-$1078597$) show some similar properties in the data, and have band gaps of $0.30$, $1.18$ and $1.73$~eV, respectively. Single self-assembled InAs quantum dots have already been demonstrated \cite{Liu2018}, and is therefore an exciting possible material to use in quantum technology. To the best of our knowledge, ZnCd$_3$Se$_4$ has yet to be synthesized and contains the toxic element Cd, which could prove an obstacle for large scale utilization. CuI, however, has recently been synthesized and was shown to exhibit remarkable optoelectronic properties \cite{Ahn2020}. Interestingly, the material exhibits a large ionic character with more similarities to other oxides in the dataset than, e.g., the more covalent materials of Si, SiC and diamond.  
It seems that ionic character alone is not an obstacle, but it is unknown at this time whether CuI has higher potential for hosting single-photon emitters, spin-manipulation or QT operation via nanostructuring. 
Note that NaCl was not predicted as suitable for QT in the insightful approach.

@@

In this work we have applied four well-known and tested machine learning methods for classification problems, these are (see for example \cite{Hastie2009,Mehta2019} for discussions and applications):
\begin{enumerate}
    \item Logistic regression,
    \item Decision trees,
    \item Random forests,
    \item Gradient boosting.
\end{enumerate}
Logistic regression \cite{Hastie2009} is a simple and frequently used method for binary and multi-category classification problems.  In addition to logistic regression, we have also applied and tested the predictions made by decision trees and ensemble methods like random forests and gradient boosting, the latter through the application of the computationally efficient  XGBoost library \cite{xgboost2016}. Gradient boosting and random forests use decision trees as weak learners and improve their predictability. For random forests this is implemented through a collection of randomized decision trees where a  subset of the features in the data sets are selected randomly when building a decision tree. Boosting methods like gradient boosting use decision trees as  weak learners and improve upon these by an iterative process that involves the estimation of the gradients of the cost/loss function \cite{Hastie2009}. Pure decision trees can easily lead to overfitting of the data under study, leading to a model that exhibits a high variance. Ensemble methods like random forests and gradient boosting on the other hand tend to soften the overfitting problem, resulting in both a small bias and a reduced variance of the model employed, see for example Ref.~\cite{Mehta2019} for an in-depth discussion of the bias-variance trade-off in machine learning. Gradient boosting implemented through the  XGBoost library, is widely used by data scientists to achieve state-of-the-art results on many machine learning challenges.

@@

The data that support the findings of this study are available from the corresponding author, and the datasets used to generate the ML models are available online at [Github link].

>>>>

%We have developed data extraction tools for numerous material science databases, and constructed over $4800$ physics-informed features for a data set consisting of more than $25000$ materials. Furthermore, we have developed and implemented three data mining approaches, termed \textit{the Ferrenti approach}, \textit{the augmented Ferrenti approach} and \textit{the insightful approach} for defining three distinct training sets for the supervised machine learning algorithms logistic regression, decision tree, random forest and gradient boost to be trained on.
Data extraction tools have been developed for a number of materials science databases resulting in a data set consisting of over $25'000$ materials. The information from the databases was converted into nearly $5'000$ physics-informed features. Furthermore, we have developed and implemented three different approaches for the data mining process: \textit{The Ferrenti approach}, \textit{the augmented Ferrenti approach}, and \textit{the intuitive approach}. 
%Rename insightful approach! To what? 
%Options include stringent, strict, cherry-picked, informed and intuitive. 
The three approaches define distinct training sets for the supervised machine learning methods logistic regression, decision trees, random forests and gradient boosting.

@@

After compiling a data set from the procedure outlined above, we face an obstacle in terms of defining a training set for the machine learning algorithms. This is challenging for two reasons. 
Firstly, although defects like the NV center in diamond and emitters in SiC are becoming
increasingly well studied, few candidate materials are known. An additional consideration
is that the physical mechanisms promoting favorable defect properties are not fully understood. Secondly, defining materials as unsuitable candidates for QT has proven to be surprisingly intricate. 
%This is not only challenging due to the lack of known suitable candidates (1),  but also due to the intricacy of defining materials as unsuitable candidates (0). 
We follow three separate procedures to define a training set consisting of materials that are labeled as both suitable and unsuitable candidates for QT.

@@

We then label \textbf{unsuitable} candidates according to the reverse requirements of the above; as materials in the ICSD database that crystallize in polar space groups, are calculated 
to be magnetic and have a band gap larger than $0.1$~eV (to exclude metals but include lower-band gap semiconductors) as calculated by DFT using the PBE functional.
The resulting complete data set contains materials where $56 \ \%$ are labeled as unsuitable and $44 \ \%$ as suitable for QT applications.

\subsubsection*{The augmented Ferrenti approach}
In the second approach, we adjust the first approach in order to improve the data set. This approach is therefore named \emph{the augmented Ferrenti approach}.

@@

For \textbf{unsuitable} candidates, we implement the same strategy as for the non-augmented Ferrenti approach. The result is a somewhat unbalanced data set, with up to $75 \ \%$ of the 
materials found in the suitable group. However, the data set is $78 \ \%$ larger than for the Ferrenti approach.

\subsubsection*{The intuitive approach}
The third approach differs substantially from the first two approaches in terms of labeling, as we apply knowledge from the field (see for instance Refs.~\cite{Atatuere2018,Zhang2020,Son2020,Toth2019,Bathen2021} for an overview)   
to pick our promising material host candidates. 
In other words, we restrict ourselves to materials where quantum compatible properties have been either experimentally demonstrated or theoretically predicted. 
We therefore name this scheme \emph{the intuitive approach}. Due to the concern of having a data set that is too narrow, we will also include some  materials that are promising in the sense that they have suitable properties for accommodating  deep defects that potentially can exhibit quantum effects - even though such effects may not yet have been shown. 
Materials where properties like single-photon emission and spin manipulation are attributed to quantum dots (formed by self-assembly or lithographic structuring) will also be included in this context.  
The approach to pick \textbf{suitable} candidates can be summed up as;   
\begin{enumerate}
    \item Pick candidates that match the formulas in Table~\ref{tab:qt-materials}, or the formulas ZnSe, AlP, GaP, AlAs, ZnTe, CdS \cite{Weber2010} or SiGe \cite{Hardy2019}. 
    \item Make sure that the candidates are present in the ICSD database.  
    \item Perform a manual screening for correct structures. 
\end{enumerate}

@@

Since the data set constituting suitable candidates in the third approach contains few entries, we add $400$ \emph{unsuitable} candidates. These are picked at random from the pool of unsuitable candidates from the two previous approaches, in addition to those that were marked as unsuitable through the manual screening process.

@@

To visualize the complexity of the training sets, we have found the two largest eigenvalues of the covariance matrix of the initial data from the Materials Project, and transformed the training sets according to the corresponding two eigenvectors. The resulting scatter plots are found in Figure~\ref{fig:2dscatterplotpca}. 
Green squares display suitable candidates and red triangles represent unsuitable candidates. 
Due to the complexity of reducing the large amount of features down to only two, both suitable and unsuitable candidates for the Ferrenti and the augmented Ferrenti approaches are largely overlapping. 
Hence, using these approaches for predicting QT material hosts could prove challenging for any model that would try to glean a clear-cut boundary between materials that are and are not suitable for QT. 
%Nonetheless, we note that the unsuitable candidates appear to be restricted to a certain parameter space while the suitable candidates extend much further.  
Therefore, we can expect that the two Ferrenti approaches would need supplementary dimensions for further distinguishing between the materials in the two categories (suitable and unsuitable), or that it could prove challenging to find a generalized model based on the available data sets.

@@

It is important to note that the constructed data sets encompass compounds formed by a plethora of combinations of surfaces, interfaces, nanostructures, compositions and structures. We note that this complexity is not necessarily reflected in the descriptors. 
%Additionally, we acknowledge that many compositions deemed as suitable candidates consist of either rare or dangerous elements. By utilizing an enormously large database as Materials Project, we have to account for their ultimate goal - to model all possible materials and their properties.
%The automated process of adding an entry to their database does not necessarily contain all relevant information about a respective material. This is information that needs to be added manually.
Furthermore, we have utilized data obtained from high-throughput density functional theory calculations. Indeed, there are possible errors associated with every step, starting from an initial calculation, adding of data in the database, gathering of data, featurization of data, preprocessing of data, data mining, and finally training a model and making a prediction. Unfortunately, if an error has happened in the first part of the process, the error follows the entire process and will get increasingly harder to detect. Therefore, we are dependent on that the Materials Project has obtained data with high quality, and we note that it is likely that there are some errors present in our data sets.

@@

Taking the four machine learning algorithms outlined above and applying them to the data sets optimized using each of the three approaches yields 12 sets of results for predicting  materials as suitable material hosts for QT. In this section, we present representative ML results based on each of the three data mining approaches.

@@

We also find two compositions with the same chemical formula, the orthorhombic coordinated (mp-$629458$) and the chalcopyrite structured BC$_2$N (mp-$1008523$) where the first structure is in a polar space group while the latter is not. The band gaps are in MP calculated as $1.85$ eV and $1.65$ eV, respectively. BC$_2$N is known as heterodiamond and is a hybrid of diamond and BN. Both structures have, as expected, strong covalent character and have been studied for application as nanostructures \cite{Gao2017}, hydrogen storage \cite{Cai2017} and superhard materials \cite{Li2017, Jiang2020} in ab-initio calculations. Of similar compounds, it has been predicted that the diamond-like structure of BC$_3$N can be a prominent spin qubit material host \cite{Wang2020SpinQB}. By creating a boron (B) vacancy, it will immediately lead to an NV center with similar properties as found in the NV center in diamond. If this is also possible for BC$_2$N, remains to be seen. We note that BC$_3$N is not present in MP, and therefore not present in our data set.

The compounds InAs (mp-$20305$), CuI (mp-$22895$ and mp-$569346$) and ZnCd$_3$Se$_4$ (mp-$1078597$) show some similar properties in the data, and have band gaps of $0.30$, $1.18$ and $1.73$~eV, respectively. Single self-assembled InAs quantum dots have already been demonstrated \cite{Liu2018}, and is therefore an exciting possible material to use in quantum technology. To the best of our knowledge, ZnCd$_3$Se$_4$ has yet to be synthesized and contains the toxic element Cd, which could prove an obstacle for large scale utilization. CuI, however, has recently been synthesized and was shown to exhibit remarkable optoelectronic properties \cite{Ahn2020}. Interestingly, the material exhibits a large ionic character with more similarities to other oxides in the data set than, e.g., the more covalent materials of Si, SiC and diamond.  
It seems that ionic character alone is not an obstacle, but it is unknown at this time whether CuI has higher potential for hosting single-photon emitters, spin-manipulation or QT operation via nanostructuring. 
Note that NaCl was not predicted as suitable for QT in the insightful approach.

@@

In this work we have applied four well-known and tested machine learning methods for classification problems, these are (see for example \cite{Hastie2009,Mehta2019} for discussions and applications):
\begin{enumerate}
    \item Logistic regression,
    \item Decision trees,
    \item Random forests,
    \item Gradient boosting.
\end{enumerate}
Logistic regression \cite{Hastie2009} is a simple and frequently used method for binary and multi-category classification problems.  In addition to logistic regression, we have also applied and tested the predictions made by decision trees and ensemble methods like random forests and gradient boosting, the latter through the application of the computationally efficient  XGBoost library \cite{xgboost2016}. Gradient boosting and random forests use decision trees as weak learners and improve their predictability. For random forests this is implemented through a collection of randomized decision trees where a  subset of the features in the data sets are selected randomly when building a decision tree. Boosting methods like gradient boosting use decision trees as  weak learners and improve upon these by an iterative process that involves the estimation of the gradients of the cost/loss function \cite{Hastie2009}. Pure decision trees can easily lead to overfitting of the data under study, leading to a model that exhibits a high variance. Ensemble methods like random forests and gradient boosting on the other hand tend to soften the overfitting problem, resulting in both a small bias and a reduced variance of the model employed, see for example Refs.~\cite{Hastie2009,Mehta2019} for an in-depth discussion of the bias-variance trade-off in machine learning. Gradient boosting implemented through the  XGBoost library \cite{xgboost2016}, is widely used by data scientists to achieve state-of-the-art results on many machine learning challenges.

@@

The data sets that support the findings of this study are available from the corresponding author, and the data sets used to generate the ML models are available online at [Github link].
